Functory: Une bibliothèque de calcul distribué \\ pour Objective Caml\footnotetext{Ce travail a été réalisé en partie dans le cadre du projet U3CAT (\emph{Unification of Critical C Code Analysis Techniques}, ANR-08-SEGI-021).}

### 2010-10-13 doc/jfla-2011/main.tex
*** empty log message ***

\newcommand{\Ocaml}{OCaml}
\newcommand{\functory}{\textsf{Functory}}
\newcommand{\JoCaml}{Jo{\&\!}Caml}
\newcommand{\unix}{\textsc{Unix}}
\newcommand{\coeur}{c\oe ur}
\newcommand{\coeurs}{c\oe urs}

@@

\section{API}\label{sec:API}
In this section, we describe the
generic API which we shall extend with several specialized
functions in the next section. The main function in our API follows the idea
sketched in the introduction. It has the following signature:
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
Tasks are pairs, of type \of{'a * 'c}, where the first component is
passed to the worker and the second component is local to the master.
The \ocaml{worker} function should be pure and is executed in parallel
in all worker processes. The function \ocaml{master}, on the
contrary, can be impure and is only executed in the master process.
The \ocaml{master} function typically stores results 
in some internal data structure.
Additionally, it may produce new tasks, as a list of type 
\of{('a * 'c) list}, which are then appended to the current set of
pending tasks.
The next section will describe the usage of this generic interface to
perform the traditional map and fold operations.

Actually, our library provides not just a single \ocaml{compute}
function as above, but instead five different versions
depending on the execution context. 
The first two contexts are the following: 
\begin{enumerate}
\item \textbf{Purely sequential execution:}
  this is mostly intended to be a reference implementation
  for performance comparisons, as well as for debugging;

\item \textbf{Several cores on the same machine:} 
  this implementation is intended to distribute the computation over a
  single machine and it makes use of \unix\ processes;
\end{enumerate}
The next three implementations are intended for distributing the
computation over a network of machines.
\begin{enumerate}
\setcounter{enumi}{2}
\item \textbf{Same executable run on master and worker machines:}
  this implementation makes use of the ability to marshal \Ocaml\
  closures and polymorphic values.
  Depending on whether the program is run as a master or as a worker,
  the relevant arguments of \ocaml{compute} are used.

\item \textbf{Master and workers are different programs, compiled with
    the same version of \Ocaml:} 
  we can no longer marshal closures but we can still
  marshal polymorphic values. As a consequence, 
  the \ocaml{compute} function is split into two
  polymorphic functions, to implement the master and workers 
  respectively:%
\vspace{-1em}
\begin{ocaml}
val Worker.compute : ('a -> 'b) -> unit -> unit
val Master.compute : 
  ('a * 'c -> 'b -> ('a * 'c) list) -> 
  ('a * 'c) list -> unit
\end{ocaml}

\item \textbf{Master and workers are different programs, not even
    compiled with the same version of \Ocaml:} we can no
  longer use marshaling, so the
  \ocaml{compute} function is split into two monomorphic functions
  over strings:%
\vspace{-1em}
\begin{ocaml}
val Worker.compute : (string -> string) -> unit -> unit
val Master.compute : 
  (string * 'c -> string -> (string * 'c) list) -> 
  (string * 'c) list -> unit
\end{ocaml}
\end{enumerate}

Our library is organized into three modules: \of{Sequential} for the
pure sequential implementation, \of{Cores} for multiple cores on the
same machine and \of{Network} for a network of machines, respectively.
The \of{Network} module itself is organized into three sub-modules,
called \of{Same}, \of{Poly} and \of{Mono},
corresponding to contexts 3, 4 and 5 above. 
The next section presents more functions provided by \functory, which
are all derived from the generic API just presented.

% \section{Derived API}\label{sec:derived}

In most cases, the easiest way to parallelize an execution it to make
use of operations over lists, where processing of the
list elements are done in parallel. 
The previous section illustrated one such case.
To facilitate such a processing,
we now derive the most commonly used list operations from our generic
API.

The most obvious operation is the traditional map operation over
lists, that is:
\begin{ocaml}
  val map : ('a -> 'b) -> 'a list -> 'b list
\end{ocaml}
The next natural operation is a combination of map and fold
operations, that is a function like
\begin{ocaml}
 val map_fold : map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 'c -> 'a list -> 'c
\end{ocaml}
which, given two functions, an accumulator \of{a} and a list \of{l}, computes
\begin{equation}\label{eq:map-fold}
  \of{fold} ... (\of{fold} (\of{fold} ~ a ~ (\of{map} ~ x_1)) (\of{map} ~ x_2))
  ... (\of{map} ~ x_n)
\end{equation}
for some permutation $[x_1,x_2,...,x_n]$ of the list \of{l}.
We assume that the \of{map} operations are always performed in parallel.
Regarding \of{fold} operations, we distinguish two cases:
\begin{itemize}
\item either \of{fold} operations are computationally less expensive
  than \of{map} and we perform them locally;
\item or \of{fold} operations are computationally expensive and we
  perform them in parallel.
\end{itemize}
We thus provide two functions \of{map_local_fold} and \of{map_remote_fold}.

In the case of \of{map_remote_fold}, only one \of{fold} operation can
be performed at a time (possibly in parallel with \of{map}
operations), as obvious from~(\ref{eq:map-fold}). However, there are
cases where several \of{fold} operations can be performed in parallel,
as early as intermediate results of \of{fold} operations are available.
This is the case when \of{fold} is an associative operation (which
implies that types \of{'b} and \of{'c} are the same).
Whenever
\of{fold} is also commutative, we can perform even more \of{fold}
operations in parallel. Thus our API provides two functions
\of{map_fold_a} and \of{map_fold_ac} for these two particular cases.
The five operations of the derived API are summarized
in Figure~\ref{fig:derived}.
\begin{figure}[t]
  \begin{ocaml}
    val map : 
      f:('a -> 'b) -> 'a list -> 'b list 
    val map_local_fold : 
      map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
      'c -> 'a list -> 'c 
    val map_remote_fold : 
      map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
      'c -> 'a list -> 'c 
    val map_fold_ac : 
      map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 
      'b -> 'a list -> 'b 
    val map_fold_a : 
      map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 
      'b -> 'a list -> 'b
  \end{ocaml}
  \caption{Derived API.}
\label{fig:derived}
\end{figure}

These five functions are actually derived from the generic API.  The
generality is achieved by implementing these derivations as a functor.
This functor, \of{Derive}, is parameterized by the \of{compute}
function, as follows:
\begin{ocaml}
module Derive
  (X : sig
     val compute : 
       worker:('a -> 'b) -> master:('a * 'c -> 'b -> ('a * 'c) list) ->
       ('a * 'c) list -> unit
   end) = struct ... end
\end{ocaml}
We now explain how each function from Figure~\ref{fig:derived} is
implemented using \of{compute} inside the functor body.

\paragraph{\of{map}.} 
Function \of{map} is easily implemented by associating an integer
index with each element (of type \of{'a}) of the input list. Thus the
list of tasks passed to \of{compute} is a list of pairs of type \of{'a * int}.
As soon as an intermediate result is available, it is passed to
\of{master}, which stores it in a local table using the index as a key
and generates no new task. Whenever \of{compute} returns, we easily
recover the output list from the values stored in the table.

\paragraph{\of{map_local_fold}.} 
Function \of{map_local_fold} is implemented using a local reference
storing the current accumulator (of type \of{'c}).  As soon as an
intermediate result is available, the \of{master} combines it with the
accumulator using the function \of{fold}. Whenever \of{compute}
returns, we simply return the current accumulator.

\paragraph{\of{map_remote_fold}.}
In this case, combining the accumulator with an intermediate result is
itself a task to be performed in parallel with \of{map} operations.
Thus tasks are of two different kinds and we use a sum type to make
the distinction.
When an intermediate \of{map} result is available, the
accumulator may be already in use in a \of{fold} operation. Hence we
need to store the intermediate \of{map} result in a table local to the
master, to be
used as soon as the accumulator becomes available.
The \of{master} function distinguishes between \of{map} and \of{fold}
results and, accordingly, updates the table and the accumulator. For a
\of{map} result, either we combine it with the accumulator in a new
task or we simply store it in the local table. For a \of{fold} result,
either we simply store it into the accumulator or we immediately
combine it with a pending \of{map} result from the table.
Whenever \of{compute} returns, the accumulator must be available and
we simply return it.

\paragraph{\of{map_fold_ac}.}
The function \of{map_fold_ac} is easier than the previous one, since
we do not need to store more than one intermediate result. Indeed, as
soon as two intermediate results are available, we immediately combine
them using a \of{fold} operation. Thus, whenever the \of{master}
receives some result, it either stores it or combines it with the
accumulator, depending on the availability/non availability of the
accumulator.  Whenever \of{compute} returns, the accumulator must be
available and we simply return it.

\paragraph{\of{map_fold_a}.}
Function \of{map_fold_a} is definitely the trickiest. Indeed, the
operation being only associative, we can only combine \emph{adjacent}
intermediate results. In order to perform this, we store intermediate
results in a local table according to indices of the input list. More
precisely, if we have already computed
\begin{displaymath}
  x_i \oplus x_{i+1} \oplus \dots \oplus x_j
\end{displaymath}
where $\oplus$ denotes the \of{fold} operation, then we associate this
value with indices $i,j$. As soon as a result of indices $k,i-1$ or
$j+1,k$ is available, it can be combined with the result of $i,j$.
Whenever \of{compute} returns, the table should only contain a result
corresponding to the full input list and we simply return it.

@@

@@

% LocalWords:  sérialisation boutisme inatteignables

>>>>

\newcommand{\Ocaml}{OCaml}
\newcommand{\functory}{\textsf{Functory}}
\newcommand{\JoCaml}{Jo{\&\!}Caml}
\newcommand{\unix}{\textsc{Unix}}
\newcommand{\oeuvre}{\oe uvre}
\newcommand{\coeur}{c\oe ur}
\newcommand{\coeurs}{c\oe urs}

@@

\section{Interface}\label{sec:API}
Cette section décrit l'interface de la bibliothèque \functory.
La fonctionnalité primitive est une fonction \of{compute} réalisant
l'idée principale évoquée dans l'introduction.
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
Les tâches sont des paires, de type \of{'a * 'c}, où la première
composante sera transmise à l'ouvrier et la seconde conservée
par le patron. La fonction \of{worker} doit être observationnellement
pure et sera exécutée en parallèle par tous les ouvriers. La fonction
\of{master}, au contraire, peut être impure et sera exécutée
uniquement au sein d'un unique processus séquentiel.
Cette fonction \of{master} accumule typiquement les résultats renvoyés
par les ouvriers dans une structure locale. Elle peut en outre
produire de nouvelles tâches, 
sous la forme d'une liste de type \of{('a * 'c) list}, 
qui sont alors ajoutées aux tâches restant à effectuer.
Le troisième argument de \of{compute} est la liste des tâches
initiales, qui déclenchent le calcul.
La fonction \of{compute} rend la main lorsque toutes les tâches ont
été effectuées. Il n'y a pas de résultat renvoyé, mais uniquement des
effets de bord de la fonction \of{master}.
Nous montrerons à la fin de cette section comment en déduire des
fonctions d'ordre supérieur usuelles telles que \of{map} ou \of{fold}.

En réalité, la bibliothèque \functory\ fournit \emph{cinq}
mises en {\oe}uvre différentes de la fonction \of{compute}, correspondant à
cinq contextes d'utilisation différents.
Les deux premiers sont les plus simples.
\begin{enumerate}
\item \textbf{Exécution purement séquentielle :}
  Elle permet d'obtenir un code de référence, pour mesurer des
  performances ou mettre au point son programme.

\item \textbf{Plusieurs \coeurs\ sur une même machine :} 
  Il s'agit là de distribuer le calcul sur une unique machine,
  uniquement en créant des processus fils.
\end{enumerate}
Les trois contextes suivants correspondent à une distribution du
calcul sur un réseau de machines.
\begin{enumerate}
\setcounter{enumi}{2}
\item \textbf{Patron et ouvriers sont matérialisés par un même exécutable :}
  Cette mise en {\oe}uvre exploite la capacité d'\Ocaml\ à sérialiser
  fonctions et valeurs polymorphes de manière portable.
  Selon que le programme est exécuté comme le patron ou comme un
  ouvrier, les arguments pertinents de la fonction \of{compute} sont utilisés.

\item \textbf{Patron et ouvriers sont matérialisés par différents
    programmes, compilés avec la même version d'\Ocaml\ :} 
  Il n'est plus possible de sérialiser des fonctions mais on peut
  encore sérialiser des valeurs polymorphes.
  En conséquence, la fonction \of{compute} est présentée sous la forme
  de deux fonctions, servant respectivement à réaliser le patron et
  les ouvriers :
\vspace{-0.5em}
\begin{ocaml}
val Worker.compute : ('a -> 'b) -> unit
val Master.compute : ('a * 'c -> 'b -> ('a * 'c) list) -> ('a * 'c) list -> unit
\end{ocaml}

\item \textbf{Patron et ouvriers sont matérialisés par différents
    programmes, qui ne sont même pas compilés avec la même version
    d'\Ocaml\ :} Il n'est plus possible d'utiliser la sérialisation
  d'\Ocaml\ et la fonction \of{compute} est présentée sous la forme
  de deux fonctions ne manipulant plus que des chaînes de caractères :%
\vspace{-0.5em}
\begin{ocaml}
val Worker.compute : (string -> string) -> unit
val Master.compute : (string * 'c -> string -> (string * 'c) list) ->
                    (string * 'c) list -> unit
\end{ocaml}
\end{enumerate}
La bibliothèque \functory\ est donc organisée en trois modules :
\of{Sequential} pour le calcul purement séquentiel ; \of{Cores} pour
le calcul distribué sur plusieurs \coeurs\ d'une même machine ; et
enfin \of{Network} pour le calcul en réseau. Ce dernier module
comporte trois sous-modules, appelés respectivement \of{Same},
\of{Poly} and \of{Mono}, correspondant aux situations 3, 4 et 5 ci-dessus.

%%% derived API %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Fonctions dérivées.}\label{sec:derived}
Dans de nombreuses situations, la mise en \oeuvre\ la plus simple du
parallélisme consiste à appliquer une opération sur une liste, le
traitement de chaque élément pouvant être réaliser en parallèle.
\functory\ fournit plusieurs fonctions d'ordre supérieur
offrant du calcul parallèle sur des listes, toutes dérivées de la fonction
\of{compute}. En particulier, elles sont disponibles dans les cinq
modules décrits ci-dessus.

L'opération la plus naturelle est celle consistant à appliquer une
fonction à tous les éléments d'une liste, c'est-à-dire
\begin{ocaml}
  val map : ('a -> 'b) -> 'a list -> 'b list
\end{ocaml}
Plus subtilement, on peut combiner une fonction \of{map:'a -> 'b}
avec une fonction \of{fold:'c -> 'b -> 'c} pour calculer, à partir d'une liste
$l$  et d'un accumulateur initial $a$, la valeur finale
\begin{equation}\label{eq:map-fold}
  \of{fold} ~ ... ~ (\of{fold} ~ (\of{fold} ~ a ~ (\of{map} ~ x_1)) ~
  (\of{map} ~ x_2)) ~ ... ~ (\of{map} ~ x_n)
\end{equation}
pour une certaine permutation non spécifiée $[x_1,x_2,...,x_n]$ 
de la liste \of{l}.
On peut alors distinguer deux cas, selon que l'opération \of{fold} est
beaucoup moins coûteuse que l'opération \of{map}, et peut être
effectuée localement par le patron, ou qu'au contraire elle peut être
coûteuse et a donc intérêt à être effectuée en parallèle des
opérations \of{map}. La bibliothèque fournit donc deux fonctions,
correspondant à ces deux cas de figure :
\begin{ocaml}
  val map_{local,remote}_fold : map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 'c -> 'a list -> 'c   
\end{ocaml}

Dans le cas de \of{map_remote_fold}, une seule opération \of{fold}
peut être effectuée à la fois (possiblement en parallèle d'opérations
\of{map}), comme le montre l'équation~(\ref{eq:map-fold}).
Il existe cependant des situations dans lesquelles plusieurs
opérations \of{fold} peuvent être effectuées en parallèle, dès que des
résultats intermédiaires de \of{map} sont disponibles. C'est le cas
notamment lorsque l'opération \of{fold} est associative (ce qui
implique que les types \of{'b} et \of{'c} sont égaux).
Lorsque l'opération \of{fold} est de plus commutative, on peut
effectuer encore plus d'opérations \of{fold} en parallèle.
Notre interface fournit donc deux autres fonctions pour ces cas
particuliers :
\begin{ocaml}
  val map_fold_ac : map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 'b -> 'a list -> 'b 
  val map_fold_a  : map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 'b -> 'a list -> 'b
\end{ocaml}

\hrulefill ICI\hrulefill

@@

% TODO : insister sur la simplicité de Same, pour les petits programmes
% notamment

@@

% LocalWords:  sérialisation boutisme inatteignables observationnellement
% LocalWords:  sous-modules
